---
title: Reproduce the `example_bmi_1` variable
author: Adrienne Stilp
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document
---

```{r setup, echo = FALSE}
options(knitr.table.format = 'html')
knitr::opts_chunk$set(message = FALSE)
```

This document explains how to use the information in the JSON documentation to reproduce a given harmonized variable using other harmonized variables.
To reproduce these types of variables, you will need to reproduce each component variable separately and then work with those harmonized values to reproduce the final harmonized variable.

In this document, we'll work with the `example_bmi_1.json` documentation.
As we'll see soon, this harmonzied variable is produced from two other harmonized variables: `example_height_1` and `example_weight_1`.
Please see the associated files for instructions on how to reproduce those harmonized variables from the original study data:

* harmonized-variable-reproduction-height.md
* harmonized-variable-reproduction-weight.md

In those files, we saved the final, combined harmonized variable in tab-separated text files with the required identifiers.
We'll use those saved files in this harmonization.



# Setup

First, we'll need to load some packages to help work with the files.

```{r setup-packages}
library(readr)
library(dplyr)
library(stringr)
library(jsonlite)
library(xml2)
```

Next, we'll read in the JSON documentation for `example_bmi_1`.
```{r read-json}
json <- fromJSON('example_bmi_1.json', simplifyDataFrame = FALSE)
```

# Reproducing the harmonized values

In this case, there is only one harmonization unit that uses the combined harmonized data:
```{r}
unit <- json$harmonization_units[[1]]
unit
```

You can see that the two required variables are `example_height_1` and `example_weight_1`.
We will read in the harmonized values for these variables that were created from the `example_height_1.json` and `example_weight_1.json` documents.
```{r}
height <- readr::read_tsv("example_height_1.tsv")
head(as.data.frame(height))


weight <- readr::read_tsv("example_weight_1.tsv")
head(as.data.frame(weight))
```

We can determine the mapping between study subject identifier and `topmed_subject_id` from these files.
We'll concatenate the identifier columns and then select the unique sets of identifiers.
```{r}
subj_map <- height %>%
  bind_rows(weight) %>%
  select(SUBJECT_ID, unique_subject_key, topmed_subject_id) %>%
  # Select unique sets.
  distinct()
head(subj_map)
tail(subj_map)
```

We should also verify that there is a one-to-one mapping between `unique_subject_key` and `topmed_subject_id`:
```{r}
stopifnot(all(!duplicated(subj_map$unique_subject_key)))
stopifnot(all(!duplicated(subj_map$topmed_subject_id)))

```
If these checks fail, you will need to verify that you used the same study mapping integers as provided in the `example-phs-mapping.tsv` file when harmonizing the height and weight component variables.
You will also need to check that you created the topmed_subject_id with a different `unique_id` for each different study accession on dbGaP.
This check is essential because the harmonization functions generally combine data values for the same subject from the different component traits using `topmed_subject_id`.

Next, create the `phen_list` data structure from these harmonized_variables.
Unlike the examples for `example_height_1` and `example_weight_1`, the `phen_list` structure will not have a `source_data` element.
Instead, it will have a `harmonized_data` element.
This element is al ist containing harmonized data, with one element for each component harmonized trait.
The name of those elements should be the name of that component variable.
The contents of the elements is a data frame with columns `topmed_subject_id`, the component harmonized variable name (e.g., `example_height_1`), and the corresponding age variable, if it exists (e.g., `age_at_example_height_1`).
```{r}
harmonized_data <- list()
harmonized_data[["example_height_1"]] <- height %>%
  select(topmed_subject_id, example_height_1, age_at_example_height_1)
harmonized_data[["example_weight_1"]] <- weight %>%
  select(topmed_subject_id, example_weight_1, age_at_example_weight_1)
phen_list <- list(harmonized_data = harmonized_data)
lapply(phen_list$harmonized_data, head)
```

Next, we can extract the harmonization function from the JSON documentation for this harmonization unit.
```{r}
cat(unit$harmonization_function)
harmonize <- eval(parse(text = unit$harmonization_function))
```

Next, we can run the harmonization on the `phen_list` structure that we just created to produce the final harmonized variable.
```{r}
out <- harmonize(phen_list)
head(out)
tail(out)
```

As before, this harmonization function returns a data frame with columns `topmed_subject_id`, the phenotype concept, and age.

Finally, we can add the subject identifiers back to this data frame, and rename variables to add the phenotype concept from the JSON documentation.
```{r}
harmonized <- out %>%
  left_join(subj_map, by = "topmed_subject_id") %>%
  select(SUBJECT_ID, unique_subject_key, topmed_subject_id, everything()) %>%
  rename(
    example_bmi_1 = example_bmi,
    age_at_example_bmi_1 = age
  )
head(as.data.frame(harmonized))
```

We'll save this harmonized variable in the example data directory.
```{r}
readr::write_tsv(harmonized, "example_bmi_1.tsv")
```

To check if you have run the harmonization correctly, please compare this file against the harmonized data in the `example_data/harmonized/example_bmi_1.tsv` file.

```{r}
check <- readr::read_tsv("example_data/harmonized/example_bmi_1.tsv")
stopifnot(all_equal(check, harmonized))
```
