---
title: An introduction to the TOPMed DCC's harmonized variable documentation
author: Adrienne Stilp
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document
---

```{r setup, echo = FALSE}
options(knitr.table.format = 'html')
knitr::opts_chunk$set(message = FALSE)
```

This repository contains documentation for phenotypes harmonized by the [Trans-Omics for Precision Medicine (TOPMed) Program](https://www.nhlbiwgs.org/)'s Data Coordinating Center (DCC).

This document gives an introduction to the [JSON](https://www.json.org/) documentation of harmonized variables provided in this repository.
Because the harmonization is done using [R](https://www.r-project.org/), we'll show how to work with the documentation using R.
All study data used for harmonization comes from [dbGaP](https://www.ncbi.nlm.nih.gov/gap/).
The TOPMed DCC's harmonization process will be described further in a paper.
The reference will be added upon publication.

## Repository contents

| file | description  |
|---|---|
| `example/`               | A directory containing reproducible examples using simulated data. See the [Examples section](#examples) for more information. |
| `harmonized-variable-documentation/` | A directory containing JSON documentation file for variables harmonized by the TOPMed DCC. Documentation files for the harmonized variables are located in dataset subdirectories within this directory. |
| `schema.json`            |  Schema definition for the JSON documentation. |
| `phs-mapping.tsv`        | A mapping between TOPMed study and dbGaP study accession. See the [Study mapping file](#study-mapping-file) section for more information.  |
| `README.md`  | This document. |
| `README.Rmd`  | Source for this document. |


## Study mapping file

The phenotypes for each TOPMed study come from one or more dbGaP study accession.
The `phs-mapping.tsv` file provides a mapping between a dbGaP study accession and the TOPMed study that its phenotypes correspond to.
It also gives an integer that can be used to generate a unique integer identifier for subjects when reproducing the harmonized variables from the documentation provided in this repository.
Including these unique identifiers in the harmonized subject identifiers is necessary because the same submitted subject identifier can appear in two different studies for different subjects.
For a single study with phenotypes located in two different study accessions on dbGaP, the unique id values (and the TOPMed study abbreviation) in this file are the same.

## Example harmonization

This directory contains simulated study dbGaP data and JSON documentation for three example harmonized variables created from the simulated study data.
There are also three example JSON documents showing how to reproduce the harmonized variables.
See `example/README.md` for instructions on running the examples.


## Harmonized variable documentation

All information necessary to reproduce a harmonized variable is available in a single JSON document file.
The `harmonized-variable-documentation` subdirectory contains these files in dataset-specific directories.

To start, we'll read a JSON document into R, using one of the reproducible example files from the `examples` subdirectory.
```{r read-json}
json <- jsonlite::fromJSON('example/example_height_1.json', simplifyDataFrame = FALSE)
```

### Metadata

The first keys provide metadata for the entire harmonized variable.

Here are keys describing standard metadata required for phenotype variables on dbGaP.
```{r json-metadata}
json[c('name', 'description', 'data_type', 'measurement_units')]
```

In some cases, we harmonize multiple variables representing the same phenotype concept.
```{r json-pheno-concept}
json[c('phenotype_concept', 'concept_variant')]
```
Different variables with the same concept have different `concept_variant` numbers.
We construct the variable name (`name`) by concatenating the `phenotype_concept` and the `concept_variant` values with an underscore.

We also provide some internal identifiers that we can use to link the documentation to the harmonized variable at the TOPMed DCC.
```{r}
json[c('dcc_harmonization_id', 'version', 'date_harmonized')]
```
The DCC can update a harmonized variable either by adding new studies or by changing how a given study's variables are processed.
When the variable is updated, the `version` given in this documentation is incremented to indicate that it has been updated, and `dcc_harmonization_id` is updated to the internal identifier of the new variable.

Last, we provide comments about harmonization in Markdown format.
```{r}
cat(json$dcc_harmonization_comments)
```
These comments give a brief description of the harmonization algorithm.
They can also include notes about the presence of a large cluster of outliers; potential differences between studies that were not important enough for removal of that study from harmonization; or a list of assays used for measuring a specific variable.

### Provenance

In addition to metadata, the JSON documentation contains the provenance of the harmonized variable.
When performing harmonization, we work in "harmonization units".
A harmonization unit is a group of subjects from a single study whose variables are processed together to produce harmonized values.
It often corresponds to a single study (e.g., the Amish or a subcohort within a study).

The `harmonization_units` key contains an array of objects, one for each harmonization unit.
Here is the first harmonization unit for this harmonized variable.
```{r json-unit}
json$harmonization_units[[1]]
```

Here are the elements that are present for each harmonization unit:

* `name`: the name of the harmonization unit
* `component_study_variables`: a list of dbGaP variable accession numbers that were used to produce the values for the subjects included in this unit. Most units contain multiple component study variables: at least one representing the phenotype itself, and at least one for the age at measurement.
* `component_harmonized_variables`: a list of harmonized variables used to produce the values for the subjects included in this unit. In this example, no component harmonized variables were used.
* `harmonization function`: the definition of the R function that converts the component variable values into harmonized variable values for the group of subjects included in this unit.

Typically, harmonization units have either component study variables or component harmonized variables, but not both.

More details about how to use this information are available in the three example documents in the `examples` directory.
See the Example Harmonization section above for more information.

## dbGaP consent considerations

We do not provide participant consent values with the harmonized phenotype documentation.
Because the harmonized data are derived from data on dbGaP, the consent type for a given participant's harmonized data is inherited from the dbGaP data used as components for the harmonization.
Users must therefore ensure that the harmonized data are used in accordance with the data use limitations for each consent group to which they have access.
